Improving the Stability of Type Soundness Proofs in Dafny

### 2023-10-11 paper/paper.tex
things

Unfortunately, as we'll see in Section~\ref{sec:eval}, this proof approach is
not stable. Anecdotally, we realized this while attemping to scale this style
of proof up to a realistic language. With enough operations and language
features, the proof resource usages begins to vary wildly between runs,
eventually leading to mysteriously failing proofs. At its core, we believe this
instability arises from all of the afformentioned unguided reasoning the solver has to do about the
code of the typechecker and evaluator,
to get from the premise of the lemma to the premises of the IH, and then
from the conclusions of the IH to the conclusion of the case.
A patchwork solution is to add guidance in the form of assertions around the
recursive calls, to spell out the intermediate steps more directly.
\jwc{Words, about why this doesn't work in the long run}

@@

\jwc{This makes the type safety proof completely generic to the implementation
of the typecheker and evaluator. You can chance the way they're implemented, add
new cases, do any number of refactors. If you fixup the compatibility and
iversion lemmas, you're good to go. As we'll see in the next section, it also
makes the proofs vary way less.}

% \begin{itemize}
%     \item We present a systematic method for constructing mechanized type safety proofs
%     in dafny, based on the concept of semantic type soundness \cite{}.

%     \item This is more modular: suppose you add division. Then you only add one inversion lemma, fix all the isSafe proofs, and the main theorem just goes through with no change.

%     \item We demonstrate empirically that, compared to other direct methods of
%     proof, our type proofs cost fewer Dafny resources on average, and have a smaller variance.
% \end{itemize}

\section{Evaluation}
\label{sec:eval}

In this section, we demonstrate empirically that our technique as described in Section~\ref{sec:sts}
does yield more stable type soundness proofs.

\subsection*{Experimental Setup}

The different typecheckers and type soundness proofs are:

\begin{itemize}
  \item (Direct):
  \item (Monadic):
  \item (Bidirectional):
  \item (Opaque-Safety):
  \item (Full):
\end{itemize}

We run the experiment by doing \texttt{dafny measure-complexity
--iterations:500 --log-format csv} on each file containing the respective
typecheckers and type soundness proofs, and parse the CSV output into a python
script which computes statistics and builds visualizations.

>>>>

Unfortunately, as we'll see in Section~\ref{sec:eval}, this proof approach is
not stable. We realized this while attemping to scale this style
of proof up to a realistic language. With enough operations and language
features, the proof resource usage varies wildly between runs,
enough that the verification will fail at random. At its core, this
instability arises from all of the afformentioned unguided reasoning the solver has to do about the
code of the typechecker and evaluator,
to get from the premise of the lemma to the premises of the IH, and then
from the conclusions of the IH to the conclusion of the case.
A patchwork solution is to add guidance in the form of assertions around the
recursive calls, to spell out the intermediate steps more directly.
\jwc{Words, about why this doesn't work in the long run}

@@

% \jwc{This makes the type safety proof completely generic to the implementation
% of the typecheker and evaluator. You can chance the way they're implemented, add
% new cases, do any number of refactors. If you fixup the compatibility and
% iversion lemmas, you're good to go. As we'll see in the next section, it also
% makes the proofs vary way less.}

\section{Evaluation}
\label{sec:eval}
In this section, we demonstrate empirically that our technique as described in Section~\ref{sec:sts}
does yield more stable type soundness proofs. As a quantative proxy for stability, we will measure
the resource usage of our proofs. If a proof is expensive, or its cost varies wildly
between runs, this can be an indicator for future proof instability \jwc{Aaron Tomb, personal communication}.
While resource usage is the most important indicator of future proof instability, we also
base our conclusions on how much verification duration --- the wall-clock time it takes for the solver
to prove the VCs --- varies between runs.

\subsection*{Experimental Setup}
We evaluate five different typecheckers and type soundness proofs for the same
language to compare their proof stability. In all cases, the evaluator is exactly the same, a definitional
interpreter as described in the previous section.

\jwc{Give these better names}

\begin{itemize}
  \item (Monadic): This is the 
  \item (Direct): 
  \item (Bidirectional):
  \item (Opaque-Safety):
  \item (Full):
\end{itemize}

We run the experiment by running the built-in \texttt{dafny measure-complexity}
command on the files containing each type soundness proof, with the flags \texttt{--iterations:250 --log-format csv}.
Each invocation of \texttt{measure-complexity} command verifies the file $250$ times, and dumps
the verification results, durations, and resource usages to a CSV file. We then parse and analyze the CSV
file with a python script, which computes the means and standard deviations of verification duration and resource usage for the solver's task of proving correctness
of the VCs in the file's main soundness theorem.
The experiments were run on a 2023 MacBook Pro with an Apple Silicon M2 processor, and 32GB memory, runnning Dafny v4.3.0, and z3 v4.12.1.
